Kalevs Krelins is a rabbi in the Peitva Shul Synagogue in Riga and at the same time chief rabbi in Lithuania and Vilna. 

He was born in Moscow and studied in Jerusalem. He served as a Rabbi in a school in Copenhagen, Denmark, in Heidelberg, Germany, and for a Young Israel community in the United States.  

In 2012, he was asked to serve as the Rabbi of the community in Riga. Together with Shimshon Daniel Isaacson, he has been appointed as the chief rabbi of Lithuania in 2016. 

Besides, he also works as the Mashgiach (supervisor) of the European Council of Kashrut (=Jewish dietary law) (EEK) in the Baltic states. 

The interview took place in the Peitva Shul Synagogue on September 28th, 2017. 

Question: How did anti-Semitism develop in Latvia after World War II?  

Kalevs Krelins: It’s hard for me to describe, because I didn’t spend the years after the war here. As far as I know, the Soviet Union had a pretty powerful control over anti-Semitism. They were able to bring it higher or lower according to the international situation. As a child, I was raised in Moscow. For us, Riga was always the top of Jewish development compared to Moscow in terms of civility, and also in terms of religious freedom. There was no religious freedom in the Soviet Union, but in terms of what they could do here compared to what they could do in Moscow and other cities, it was an example of a very well functioning community. They had kosher meals, they baked matzah and sent it to the whole Soviet Union. They always performed circumcisions. I do circumcisions myself. I travel to other countries and people say: “We had a circumcision thirty or forty years ago, and the guy came from Riga”. Riga was always the place. For sure, there was the KGB, people were tortured. But if we compare it with Moscow, it was not so bad, and many very knowledgeable people were raised here after the war in Soviet times. That’s what I know from stories. My own experience here is limited to the last five years. I don’t see major issues with anti-Semitism here.  We all know that we cannot judge any place according to strange local people. I believe that the government here is very nationalistic, but in my opinion, their nationalism is related to the conflict between Russians and Latvians, so the Jews are not in the picture. We used to say that we can’t relax, because they’ll finish the conflict with the Russians and then they come after us, but we thank G-d that we don’t have any problem. There is a joke: they finish with them, they’ll start with us.  

Question: Do you think there is some reality behind this joke?  

Kalevs Krelins: Historically, yes. History shows that Jews were always seen as an enemy. When a new enemy came, the Jews went to the second line. As soon as there is peace with that enemy, they go back to see the Jews as their enemies. I hope I exaggerate. But we have to be aware. 

Question: In Germany, synagogues are protected by the police. Is it the same in Riga? 

Kalevs Krelins: Here we don’t talk about protection. I know how the synagogues in Germany are protected. When I was the rabbi in Heidelberg, I once parked at my parking place next to the synagogue. It was September 11th, the year after the attack. The police didn’t recognize me, so they jumped on me and checked me. They take it very seriously. Here people go around the whole country, they feel that nothing happens. It’s not only about synagogues, in the whole country, they're relaxed. They’re only afraid that Russia will invade. In Germany and France, synagogues are much more protected. I personally never had any bad feeling in Germany, but I was in the most intelligent place, Heidelberg, which is a university city. Here in Riga, I don’t feel anti-Semitism.  

Question: So you don’t have any problems walking around with a kippah? 

Kalevs Krelins: First of all, I don't walk around with my kippah. I wear a casquette, for a simple reason. I came here from New York, where it’s easy to walk around with my kippah. I know that there are many people who hate Jews, but if someone who hates Jews and wants to express himself to a Jew meets me with my kippah in New York, he can go further down the street and meet someone else with a kippah. Here in Riga, the guy who wants to say or do something has to wait a couple of years to meet the next Jew. That’s why they hurry up to express themselves. Sometimes on Saturday, I go with a hat, sometimes I go with my kippa, my kids too, but we've never had any issues. Maybe once or twice somebody said something. 

Question: How was the situation while you were growing up in Moscow? 

Kalevs Krelins: I grew up in Moscow till the age of 19. The Moscow of today and the Moscow of those days, it's a different world. There was a lot of official anti-Semitism. I went to a school for mathematically gifted kids. We were 38 kids in my class, 30 were Jews. And almost no one went to the university entrance exams, we went to lower-level colleges. Some Jews tried and they were gifted in math, but the chances of getting accepted were very low. The government controlled everything, and they maintained the level of anti-Semitism that they needed. When they had negotiations with the West, with Reagan, they lowered the level, when they wanted to push the Jews, they raised it. In the nineties, they lost control and all the things went crazy. Now the control is back in Russia more or less. We may blame them for whatever we want, but Anti-Semitism is not in the picture today. 

Question: The Russian government nowadays is not anti-Semitic at all? 

Kalevs Krelins: Very extremely not anti-Semitic. I don’t think that there was ever in the Russian history a more philo-Semitic leader than the one who is ruling now.  

Question: Why is that? 

Kalevs Krelins: I heard from many people that it is very personal. I can bring an example: Vladimir Putin is coming to some city somewhere in Russia. They show him the town, the buildings, the church, the mosque. He says: show me the synagogue. They show a small synagogue. He says: I don’t like it, rebuild it. In half a year, they build a high-scale synagogue. I mean, it's something personal. I heard he had good experiences with Jewish teachers and Jewish neighbours. Thank G-d. I’m not talking about his politics, and not even the politics with Israel, Hamas or Iran. I’m talking about Jews as a nation, there are very positive relationships. I don’t want to be blamed for praising him, I’m talking about this specific issue. And we always used to appreciate what we get. There might be some negative things, but we have appreciation. From the experience of Russian history there has never been such a positive leader for Jews.    

Question: And the general population? 

Kalevs Krelins: Different. They also have an issue with the Muslim population of Russia. They have to prioritize. 

Question: Did you experience anti-Semitism personally on the streets, not from the state?  

Kalevs Krelins: Very little. At school, maybe.  

Question: What happened?  

Kalevs Krelins: Some were making fun of the Jewish last names. It was at elementary school. When I moved to the math-gifted school, there were mostly Jewish people. No one would take a risk, not even teachers. 

Question: What are the differences regarding anti-Semitism between Germany and Latvia?  

Kalevs Krelins: My personal experience is that Germany recognized and experienced the pain of what it has done, but Latvia never did. Latvia and the Latvian people were extremely cruel with the Jews. They say that the German government gave them free rein and the Latvians did the whole job. The Latvians, Lithuanians and Ukrainians were the most cruel, more cruel than the SS themselves. The people who remember say that the German government was certainly terrible, but as people, Latvians, Lithuanians and Ukrainians were much worse than the Germans.  

Question: Is the Holocaust remembered and spoken about in Latvia? 

Kalevs Krelins: No, they say that there was the occupation. They always say that they were under occupation, we’re under German occupation, we’re under Russian occupation, we’re under Swedish occupation 400 years ago, they are always under occupation, it's not their fault. But they did their best to destroy all the Jews. The terrible thing is that the number one or number two in the destructions was Herberts Cukurs, who is a national hero. He's not oficially recognized, but officially praised for his achievements: He built planes, he flew to Arctica. There was a musical about his greatness here. We see that he has achievements, but these achivements can’t cover up what he did. He was killed by the Mossad during the sixties in Uruguay. He fled from America, from Russia, he fled from everyone. Many war criminals fled to Argentina. The Mossad killed them. They now say that it was a mistake that they didn't bring him to court. 

Question: Do you mean that Latvians are not remembering and recognizing their own guilt? 

Kalevs Krelins: No, they are not recognizing their own guilt. They do say that unfortunately there were some Latvian people who collaborated. They have Waffen-SS marching here, you know, every 16th of March. They say it's not exactly against Jews, but against Russians. When they say it's against Russians, everyone is forgiven. 

Question: Is the Jewish community offended by these marches?   

Kalevs Krelins: Definitely. The Jewish community is very offended by these celebrations. We know that some of these people from the Waffen-SS participated in mass murder.   

 

Question: Do you think that the government is distancing itself enough from this march?  

Kalevs Krelins: It’s very hard to follow what the government is doing here. I don’t want to speculate, but certain members of the parliament and of the government are participating in or praising the celebrations. And there are very few demonstrations against it, just expressions. Usually when they catch somebody who expresses himself against it, they say it is an agent of the Kremlin.  For example, when somebody screams “You are nazis”, “You are killers”, they take him aside for disturbing the public order and say it’s inspired by Putin.  

Question: Does the Jewish community organize any demonstrations against the celebrations? 

Kalevs Krelins: No, I don’t think so.  

Question: Would they be seen as agents of the Kremlin if they organized demonstrations? 

Kalevs Krelins: Yes. Latvian Jews were almost all killed during the war, ninety percent. After the war, Latvia became a very developed part of the Soviet Union. It was the Silicon Valley of the Soviet Union. I remember that all sorts of computer parts and cars were made here. Latvia was top technologically. So they sent many professionals from Russia here, among them many Jews. A big part of the Jewish population of Latvia today are descendants of Russian Jews that came as professionals to develop the economy of Latvia. So, most Latvian Jews are actually not Latvian. For us, if a person is a Jew, it doesn’t matter where he comes from. But for them, Latvian Jews and Russian Jews are different. Latvian Jews were exterminated and now these Russian Jews are associated with Russia.  

Question: Do you think that those who are anti-Russian are also anti-Semites and vice versa? 

Kalevs Krelins: A bit. I have to admit that many Jews who were oppressed in Russia went to the communist movement after the Revolution. Jews were one of the most oppresed nations, so they went against the system and joined the reds, the communist party. 

Question: Did they consider themselves as Jews? 

Kalevs Krelins: They considered themselves as Jews. They were secular, they considered themselves as people who stand against different types of oppression. And oppression of the Jews was one of the oppressions in Russia before the Revolution.   

Question: How big is the Jewish community here? 

Kalevs Krelins: We don’t have official memberships like in Germany, for example. In Germany it works with Kirchensteuer, the register. Here we don’t have such numbers. Very approximately, we talk about 8000 Jews in Latvia. I know that the Jewish community in Riga has 100 members, but that doesn’t mean that 100 people are coming to the services. On a daily basis, about 15 people come. On Saturday, we talk about 40 people more or less, and on high events it’s packed. It depends on the weather, the time of the year, but it’s a lot, 150-200, sometimes more.  

Question: How have the numbers developed since the independence of Latvia?  

Kalevs Krelins: Many people moved to Israel after the Soviet Union broke apart. The numbers have certainly gone down. But the whole population of Latvia is going down.  

Question: Was it illegal to emigrate before independence? 

Kalevs Krelins: No, it was legal, but the government did not appreciate it. People applied and they got a permission, but only a few. In the begining of the seventies, many people got the permission. After the war in Afghanistan in the early eighties until 1987, it was very hard to get a permission. After 1987, it got easier again. 

Question: Is there a discussion in the Jewish community about moving to Israel? 

Kalevs Krelins: I think it's alredy settled. Some people come to me and ask why I don't move to Israel. I think if people are here, they have a reason to be here. Young people go to Israel to learn, to go to school or college. There are many people who lived in Israel, experienced Israel, and then came back. They decide to settle here. So more or less everyone came to his or her own conclusion to this question until now. The same is happening in Germany.  

Question: Do you see any future for the Jewish community here in Riga? 

Kalevs Krelins: When someone from the Jewish community asks such a question, I have to say it’s up to you. It’s like a free market. If there is a demand, we will supply the demand. Sometimes there is a rabbi who wants to have a synagogue, so he creates himself the crowd. I think this concept is wrong. If the people need it, if they show that they want to know, they want to learn, I want to teach, but the first step has to be done by the people. Otherwise it’s counterproductive.  

Question: How are the relations between the Jewish community and the other religions? 

Kalevs Krelins: There are official meetings, sometimes I have to go and say hello, or they come to us and say hallo. On the personal level, some people are good friends, regardless of their religion or the nation.  

Question: Are there many family or friendly relationships between Jewish and non-Jewish Latvians, intermarriages for example? 

Kalevs Krelins: More Russian and Jewish, not Latvian and Jewish. This country is also divided into nations. There are Russians and Latvians. Russian-Jewish intercultural relationships are more common than Latvian-Jewish relationships.    

Question: Are the stereotypes about Jews still alive? 

Kalevs Kralins: That they crucified Jesus? 

Question: For example.  

Kalevs Krelins: I don’t know, it's hard to say.  

Question: What about the stereotype that Jews were collaboraters during Soviet times, that they were sending Latvians to Siberia? It is the most common stereotype about the Jews in Latvia.  When you try to explain that it is not like this, they say that you are a victim of Russian propaganda.  

Kalevs Krelins: Yes, but listen, we have to admit that many Jews were in high communist offices. Not actively Jewish, I mean, praying, going to Synagogue, no. They were of Jewish origin. So the stereotypes still exist, because maybe the name of the officer who was in charge of the deportations was Jewish. But it was not a national issue at all. People were not acting as Latvians, people were not acting as Jews, people were acting like bandits. Stalin killed everyone. He destroyed all the nations in the Soviet Union. The most terrible thing he did was directed against his own nation, the Georgians. There certainly were measures directed against Latvians as a nation, but also against everyone else. The aim of what happened in Germany was the final solution for the Jewish nation. In the Soviet Union, there was no final solution for any nation. They just destroyed people to impose a rule of terror in the country, to keep people under pressure regardless of their nationality. 

[We substantially modified and shortened this passage in order to make it more easily understandable. Throughout the interview, there were some difficulties with the language since English is not the mother tongue of Kalevs Krelins. We had the feeling that this passage, in its original wording, did not quite convey the argument that Mr. Krelin actually wanted to make. The full original transcript of the interview is available on demand.] 

 

  

 

 

 

 

 

 

 

 

 

 

 