\section{Dr Piotr Setkiewicz}

\textit{Piotr Setkiewicz (born in 1963) is the director of the Centre for Research at the Auschwitz-Birkenau State Museum. He graduated from the Faculty of History at the Jagiellonian University in Kraków. Mr. Setkiewicz received his doctorate in 1999 at the University of Silesia in Katowice with a thesis entitled ``IG Farben - Werk Auschwitz 1941-1945''. He is the editor-in-chief of the scientific publication The Auschwitz Journals} (Zeszyty Oświęcimskie) \textit{as the head historian at the Auschwitz Museum.\\
We met him at Auschwitz Museum and Memorial on January 27th, 2018.}\par 
\vspace*{2em}
\textbf{Have you had any first-hand or second-hand experience with anti-Semitism over the course of your life?}

\textbf{Piotr Setkiewicz:} Over the course of my life? Always, almost every day. Just yesterday, I was called by a man who is anti-Semitic, who is denying the truth, who said that I can be blamed for all mistakes and all the false information that appeared in the media about the Holocaust. He said hat I'm supporting the fake version of history which is based on Jewish, Polish, and communist lies about the Holocaust, that we are promoting the survivors who are lying, that we falsified the documents in my archive, and that we are taking money from Jews in Israeli shekels. Many such things. Many people in different countries, not only in Germany, but also in the United States, practically all over Europe, including Poland, surprisingly, try to persuade everyone that the Holocaust never happened, that Auschwitz or at least the most important objects here, like the crematoria and the gas chambers, were built only after the war, and that the gas chambers were actually used for delousing purposes, for disinfection.\\ 
My reaction usually is that I don't want to be involved in any sort of discussions with Nazis because as far as I know from my own experience, it's just a waste of my time, and the arguments always stay the same - for instance, that conditions in Auschwitz were relatively good, that it was necessary to isolate the enemies of the \textit{Reich} because it was a war and that, of course, sometimes it might happen that somebody died here. But, nevertheless, ``the number of victims is highly exaggerated, the number of one million people who were murdered in Auschwitz, and obviously, it's not true because the capacity of the crematoria was many times lower than we read in books, read in documents'', etc.\\ 
Even some strange and surprising arguments are in the discussions: For instance, there is a picture taken by somebody from the staff of \textit{IG Farben} works. \textit{IG Farben} was a huge factory of synthetic rubber; it was situated on the other side of the city. During the war, many thousand prisoners worked at the construction site there. Apart from the prisoners, of course, there was German staff, men and women, about 7,000 people, and they resided in the \textit{Siedlung}, the settlement. They lived somewhat apart from the concentration camp and they got, for instance, a sports club. There is a picture of two people who are fencing and there is a group of people in the background of the picture, and over them, there is an inscription, the name of the sports club in German, \textit{IG Auschwitz}. Now, the Nazis believe that this is a typical representation of the living conditions in Auschwitz, that the people in the picture are prisoners.\\ 
Other than that, one of the most important arguments is the presence of the swimming pool in Auschwitz, which is visible just behind us. We know that it was one of many pools or rather tanks for water for the fire brigade. In 1944, the Allies began to appear over Auschwitz and the SS suspected the threat of mass bombings and a fire, particularly in Auschwitz-Birkenau, so they built a number of such pools with reserves of water for the camp fire brigade. And surprisingly enough, in Auschwitz, the SS or rather prisoners themselves, the functionaries, the kapos or block overseers, mostly the German habitual criminal prisoners, they organised something like a competitive swimming in this pool. There was something like that, as we know from the testimonies of survivors and from somebody who was taking pictures, who was a fan. Nevertheless, there was a difference between the situation of the functionary prisoners and the real fate of regular prisoners at the concentration camp. \\
As I said, there a many such arguments, many stupid discussions. The most difficult and rather new problem is the huge amount of such articles, comments, entries on the internet. On the internet, you can write anything and there are plenty of web pages explaining that Auschwitz never happened. The problem is that recently, we are faced with more and more people denying the Holocaust in Poland which, for me, was something I could not believe because many Poles lost their lives in Auschwitz, as well as the Polish Jews, and the knowledge about the German crimes in Auschwitz was obvious for everyone in Poland, for many years after the war. Now, I am afraid that there is a third generation of people who have no personal experience. They probably heard, ``my great-grandfather was in Auschwitz'', but the distance to history is too long. Just a few days ago, the Polish television showed a ceremony of the anniversary of Adolf Hitler's birthday on the 20th of April, a group of Neo-Nazis meeting in the forest somewhere in Silesia. They were celebrating the anniversary, they had the \textit{Wehrmacht} uniforms, they baked a cake for the \textit{Führer} with the swastika.\\ 
Perhaps that is still marginal in Poland. Nevertheless, these views are shared by the people who recently protested against immigrants, against attempts from the European Union to persuade the Polish government to accept a certain amount of people from Syria and other countries. It's still not the same as with the Nazis, nevertheless, within the relatively large group of Poles who don't want to even hear about immigrants from Arab countries, there is a narrow margin of people, particularly young people – I’d say 20-25 years old, football fans, something like that - who try to make a link between their anti-immigrant views and the Nazi ideology. In the Museum, there are very rare cases where we can meet such people here. It's not a problem for us, but of course, we believe that one day, we will face the situation that a group of people with the Nazi insignia will wish to visit the Museum - so, the problem is, what to do then? Our security guards should do something and prevent it, but something like this might happen.\\
That's a good reason for us to do our work, to publish books. I don’t want to be involved in any sort of direct discussions with the Nazis. Nevertheless, we are providing arguments for many teachers, arming them with the documents. German documents, I'd say, are more useful in such discussions between students and teachers because if we gather the testimony of a survivor, it’s easier for a denier to say: ``Well, it’s just nonsense what this man is talking about''. For example, a few weeks ago I read a critical review of a book written by a certain survivor on the internet,. The Survivor wrote a book with about 300 pages about his experience in Auschwitz. This is basically a good witness - I mean, there are many survivors that are simply trying to tell us their stories. Nevertheless, there are unfortunately, and it's quite normal and natural, some people who believe that they should add something to their stories. For instance, if they were in Auschwitz, they testify about certain happenings that took place in Birkenau. It’s not true because they could have heard about the \textit{Sonderkommando} rebellion, but they had no chance to bear witness of this uprising. Nevertheless, I understand that something like this might happen, that particularly the survivors who used to take part in certain ceremonies and to meet with schoolchildren, being asked about the war experience, wish to say:``I was in Auschwitz, I saw everything''. Going back to this book: Although on one side he was generally credible, at one point in his book he said that he remembers that there was a small room in one of the barracks where the Germans, the capos, collected the bodies of those prisoners who died during the nights, and that it was only after three of four days when they collected the bodies and put them in a truck, and the truck was taking the bodies to the crematoria to be burned.\\
The author of this article on the internet made use of the fact that arguments like that are obviously not true because he found a German document in which the commander of the Birkenau ordered to remove the bodies every day. For him, this is an obvious argument that this man was lying, and if he was lying on this particular point, on this particular page of the 300-pages book, probably the remaining 300 pages are also not credible.\\
Another category of sources, of course, are the testimonies by the guards themselves. They testified before the military tribunals of the Allied powers in West Germany, immediately after the war before the Polish courts, and then in Germany in the Frankfurt trials, and in many other locations. At least some of them they accepted their guilt and admitted: ``Yes, I saw the people being selected on the ramp. I saw the people being driven to the gas chambers, I saw the thousands of bodies and the flames over the chimneys of crematoria.'' So, what is an argument of deniers against it? That they were forced to testify in this way, beaten by the guards, by British, Polish, whatever, communist, Jewish guards. That was why they had to give these false descriptions. \\
Then there is the last category - the German documents: The problem is that most of the German files from the chancelleries, from the office administration of Auschwitz, have been burned immediately before the final evacuation of the camp. If you take any testimony of a survivor who was here in January ‘45, typically you can find therein information like: ``I saw the piles of papers, of documents on the crossroads of the camp being burned by the guards''. And that's true. We've got a very small amount of original documents from German administration, a relatively large amount of documents from the German construction office, the \textit{Bauleitung}, probably because that was a separate unit of the SS-administration, which was situated in several barracks at a distance about 300 metres from here, not inside the camp. By chance, for some reason, when the SS escaped from Auschwitz, the just forgot to burn these files. \\
If you take into account all these surviving documents, they are good material for historical research, just these plans and construction files of the central construction office of the SS. The amount is about 150,000 pages of documents, a large number of documents, and they are a very valuable source, which is very useful for our indirect discussions with these deniers. There is, for instance, a well-known author who wrote a number of books and in one of them, he tried to summarise the most important arguments of the deniers. One of them was that if we talk about the first temporary gas chambers in Birkenau called ``the little red house'' and ``the little white house'', the ruins of little white house are still in Birkenau, but nevertheless he said that it’s impossible that in so many surviving documents there is not any hint about these temporary gas chambers. So, the prisoners were lying, survivors were lying, because they were Communists, Jews, Poles, whatever – that’s obvious, and the main proof that these provisional gas chambers did not exist at all is that there is nothing written in the documents.\\ 
His argument is as follows: That if something like that had existed in Birkenau, then it must have been in the records of the trips of the trucks, especially for delivering bricks and other construction materials to the site of these gas chambers. There must be something, information about the bills, about the amount of money spent for the construction of the gas chambers – but there is nothing. Thus, he tried to persuade his readers. But unfortunately, we have found seven such documents that clearly indicate that these gas chambers existed. That one of them costed 14,000 \textit{Reichsmark}, for instance, that the SS ordered three wooden barracks for each of these gas chambers, which were used as a temporary store of clothes of the victims. Each of these barracks costed about 17,000 Reichsmark and they were removed from the area in the moment when the Germans completed more modern and huge crematories in Birkenau, so the old gas chambers were dismantled and destroyed. We have found requests from the Bauleitung asking for the remaining wooden barracks, because they might be used for some other purposes in the camp. And we’ve got the information that there is a need to send 500 prisoners on a day to dig trenches around the bunkers. That was because one day in July 1942, Heinrich Himmler, during his visit in Auschwitz, ordered to empty the old mass graves and burn the bodies on piles of woods.\\ 
Then, what else have we got to prove? Apart from testimonies of survivors and apart from reports of Polish resistance members, there is a document in which the SS ordered a long electric cable to provide electricity to nineteen burning places. It was necessary to install lamps there because the burning took place not only in the day, but also during the nights. There are many direct or indirect proofs that these burning pits existed. I don’t think that you are familiar with the arguments by deniers, but we’ve got a relatively large number of counterarguments. The problem is that it’s difficult to discuss with such people because they say that all let's say 10,000 witnesses were lying and that these thousands of documents were fabricated by Poles, Jews, and communists after the war. In such circumstances, it’s rather difficult to raise any arguments.

\textbf{What do you think, what makes people want to deny the fact of Auschwitz or the Shoah?}

\textbf{Piotr Setkiewicz:} Why? For many reasons. Perhaps some of them believe that this form of government in Germany in the 30s was ideal, that everyone knew his place in the society - ``but now, what do we see around us? The politicians, thinking only about how to gain more money, who are no true patriots. And the \textit{Führer}, that was a man who solved the problem of the economic crisis in Germany. He built the highways. He restored the pride of the German nation.'' These might be the arguments of the German deniers. In other countries, I think, the reason is mainly anti-Semitism because the Holocaust, as a part of the Jewish history, is identified by most Jewish historians as a marking point in the history of the modern nation. In some books written by Jewish historians, you may see that they divided Jewish history in two periods: Before and after the Holocaust. It’s a very important point on the map of Jewish history. So, for those people who are anti-Semitic, who don’t like the modern state of Israel, the Holocaust is something which is a guilt and a giant point which should be attacked. And in the history of the Holocaust, Auschwitz is as central point. There is the famous quotation from the interview with David Irving, the leading British denier, who said: ``If we sunk the battleship of Auschwitz, it would solve all the problems of the 20th century’s history''. Auschwitz is a focus, the most important part of the story. ``If we were able to prove that Auschwitz never happened it would be easier to destroy the myth of six million of Jewish victims.''

\textbf{Would you say that these Holocaust deniers are becoming more or rather less over the last twenty-five years?}

\textbf{Piotr Setkiewicz:} Immediately after the war, it was the group of the German veterans mostly, particularly from the unions of the SS, who were deniers. There was a number of guards from Auschwitz even who took part in certain ceremonies where they solemnly declared that they were in Auschwitz, but never heard about any gas chambers or crematoria. That was the first wave of denial. Then in the 60s, there was a rise of interest in the Holocaust. On the one hand, we got the trial of Eichmann in Jerusalem, and on the other hand, a few years later, a series of so-called Auschwitz trials in Frankfurt. It was the time of student’s rebellion in Europe, the leftists began to be interested in he history of their own country. In the leftist press appeared the pictures of leading Nazis who were still living in Germany. The titles of the articles were as follows: ``The assassins, the killers are still among us''. So, there was this popular belief of the leftist people that something must be done in order to cope with our past and to punish the people who, in many cases, continued their prewar career after the war.\\ 
All those with a Nazi past whom the judges prosecuted – the technicians, engineers from the companies like \textit{IG Farben} – were renowned chemists and board directors in different companies in Germany. The young people at this time thought that we should do something with this problem of the history of their grandparents. If you observe the rise of the interest in the Holocaust, it was a ``natural'' reaction from the side of the right-wing radicals to deny the Holocaust.\\
Then in the 70s, they got some financial problems, some newspapers and journals published by the Nazi organisations disappeared. Now, I believe, we can observe again that the wave is going up – perhaps because of the internet. If in the 70s, there was a problem with publishing the journals, then now, in the era of internet, its costs almost nothing.\\ 
There is perhaps also a reaction to the idea of a united Europe. As for Brussels, we should rather think of ourselves as Europeans, to a certain extent, not only as Poles, Czechs, whatever, but we are also Europeans, members of a European family, who share the same values and the same budget. So, there is a reaction of the people who don’t like this concept of a united Europe, who think: ``Primarily, we are Poles. And we don’t want to be governed from Brussels, we got our own government, our own Polish Złoty, our own history.'' Of course, with the problem of immigrants and the terrorist attacks in other countries, among Poles there is a tendency to talk about it in this way. ``Because our friends – England, Germany – accepted many thousand Arab refugees and among them, there are of course terrorists, they got a problem. We don’t want to have any refugees on the Polish soil, so that we haven’t got the terrorist attacks and bombs in Polish cities. Therefore, we are right.''

\textbf{How is this connected to anti-Semitism?} 

\textbf{Piotr Setkiewicz:} If you believe that we are the better ones, that our values are better because we are Christians, or Europeans, it’s just one step to not liking the Jews because the Jews, ``they got too much money, they are directors of different banks, they control the world bank system, and the Jews are primarily leftists, they promote the leftist values, and because we are nationalist to a certain extent, we don’t like to accept these foreign values. There is Mr Soros, who is a Jewish billionaire, and who is supporting the leftist organisations in Poland – we don’t like him because he is leftist, and because he is Jewish.'' It is not far from this concept of national unity:``Because we are united, we represent the united nation, all Poles are brothers, and when we see these poor waves of Ukrainians, coming to Poland to work in different jobs on the construction sites, in the shops...'' Unfortunately, there is – quite human, I’d say – a tendency to look at them as people who are not us, of course, and who represent some lower cultural standards. Surprisingly, when Poles are going to the Western countries to work, they hear these anti-Polish jokes. There are very bad anti-Polish jokes, aren’t there? It is not nice to listen to such things, but now if we are looking at these Ukrainians who come to Poland, at least some Poles unfortunately believe: ``Now, we are in the positions of the British, we can look at these Ukrainians from a certain higher level.'' \\
It’s difficult – there is the question of policy, the question of a lack of education. Such tendencies to look at other countries as an enemy, as the potential threat, are a result of the modern nationalist eruptive at the end of the 19th century. There is a model of patriotic education in the schools in many countries,  to cultivate remembrance of national heroes. In many countries, there is a tendency to talk about the history of these countries in such a way: That we were always good guys, that we are always attacked by the neighbours. ``We’ve got our heroes of course, nevertheless, we are still fighting for our independence, and it is necessary to remember that independence is the most important value. We, as nation, have to remember that ourselves in such a way.'' And there is also this looking at the strangers in a specific way, that we can accept a limited number of foreigners, but if they ever exceed a certain amount, that is a potential threat for our identity. ``We are Christians, we are not Muslims.''\\
Nevertheless, Auschwitz is still present somewhere in the background of such discussions as a good illustration of what might happen if we forgot about universal values. When we see the strangers not only as the other people, but also as the representatives of the black forces of international conspiracies. And then it’s very easy to cross the line, to be on the black side of the force. There is the question of reactions to the Nazism of German society in the 30s, here in the camp, the reaction of certain people who were not Germans, but who accepted the posts of \textit{Blockältester}, the Kapos – many Poles and even Jews here in Auschwitz who received a stick and tried to exercise force over prisoners, to beat them, because the prisoners could not work with efficiency, the stick is the best argument.\\ 
These were symbols and pieces of history that might have been used in the constant discussions about the condition of the modern world and the societies here. What should we do when we hear about certain atrocities, like a genocide in other countries? Is it possible to do such a thing? We heard about the massacres in Syria, or in Iraq, or in Sudan and such countries. What should we do? There’s not a single good answer for it. Nevertheless, if we talk about it in terms of constant discussions about human rights, for instance about the need to combat racism in many countries, Auschwitz is always somewhere, is being used as a symbol, as a turning point, as a milestone in such discussions. I was surprised when I saw once a textbook for history in Singaporean schools. There was a quite comprehensive chapter about Auschwitz, on the other side of the globe. If you talk about the history of Europe and the Second World War, with, say, Koreans or people from Argentine, most people say that they know very little about it. But in 99\% of all cases, they have heard the name of Auschwitz.
 
\textbf{Do you think that Holocaust denial is the most dangerous form of anti-Semitism nowadays?}

\textbf{Piotr Setkiewicz:} This is a part of the ideology - important, but not the only one. Of course, anti-Semitism is something much wider, it’s a popular belief that the Jews are controlling the world economy, that, in some cases, leading politicians in certain countries have Jewish roots. that the governments are under the constant pressure from the national Jewish organisations. Holocaust denial is only a part of the ideology, but nevertheless, it is a very important one because anti-Semites believe that Jews are using the Holocaust as an argument to defend themselves, to blame other nations, that they wish to control the market in countries like in Poland, to get back their property that was confiscated during the war by the Nazis and which was taken over by the Communist government in these countries. That’s a good illustration of the situation in Poland.\\ Taken altogether, ``the Jews are members of the world conspiracy, and they are trying to use the argument of the Holocaust in their propaganda. We are poor guys who were persecuted over centuries and it’s virtually impossible that we can represent a threat for the others. We were the victims, not the oppressors. So if we, the deniers, destroyed this argument that the Holocaust happened, it would be much easier to preserve our arguments against Jews.'' Therefore, the history of Holocaust occupied an important position in the ideology of anti-Semites, including the ones in Arab countries – you can buy ``\textit{Mein Kampf}'' translated into Arab very easily in the larger cities in the Arab world.\\ 
These people believe these claims by Hitler that Jews are insects, that something must be done in order to remove this constant Jewish threat to the German nation, and so on. Is it possible to find an example of an official institution in Arab countries that combats such crimes by Nazis? No! I never heard about documentary initiatives in Egypt or other Arab countries against deniers, against the movies, or the internet, YouTube and Facebook activists. There is the question of politics, of course, and the question of the legacy of the Arab-Israeli wars, etc. In Europe, the situation is different. Nevertheless, the result is sometimes the same: the anti-Jewish propaganda, which is in most cases combined with fighting with the history of the Holocaust.\\ 
If you ever have the opportunity to go to my archive and to the museum for a couple of days, I’m proabably able to persuade the most stupid denier that Auschwitz happened. It would require some work and some time, but nevertheless, that’s possible. I believe that we have enough arguments and enough documents. The problem is that of course, there is nothing like the one single document, the order written by Hitler saying: ``Today, I decided to solve the Jewish problem in Europe and to kill all Jews. Adolf Hitler.'' Such a document does not exist. But after carefully analysing, say, 200 documents, even the most stupid denier would be persuaded by the force of the arguments. Even assuming that the documents were falsified, they would have to recognise that it could not be done in such a way. \\
There is, for instance, a document in which the commander of a certain unit in Auschwitz tried to give his opinion about his sergeant, promoting him before the high military ranks – some guards received military medals for their services in Auschwitz. He explained that the sergeant was very active and very enthusiastic in collecting clothes during the action of evacuation of Jews. This means that in the result of this action of evacuation of Jews, the Jews were losing their clothes. If we look at the orders for the drivers of the trucks, the trucks were bringing Zyklon-B from the factory in Erfurt. The problem was that Zyklon-B was used in Germany on a wide scale for delousing purposes, for killing insects. In one order we’ve got a sentence saying that the driver is going to Erfurt and the reason of the trip is delivery of Zyklon-B. Okay, why not. Somebody might say that Zyklon-B was used here, and that is the truth, for delousing of clothes. But there is a second order in which the reason of the trip is that the driver should bring the ``material for special Jew treatment'' or ``\textit{Materialien für die Judenumsiedlung}'', materials for resettlement of Jews. Assuming that everything was okay with the Zyklon and that the Zyklon was always used for delousing purposes here – so why, in other documents, did the German administration use such complicated euphemisms like ``materials for special treatment''? Why they did not simply name this as Zyklon-B, or as material for disinfection? My answer is that this was because this particular clerk who wrote the document that knew that actually a large quantity of Zyklon-B was used not for delousing purposes in Auschwitz, but for killing people. In this way, he wished to hide the truth. \\
So, if you give me, say, ten hours, I’d be able to persuade the strongest denier, I believe.